Verification of Stencil Computations with Spatial Specifications

### 2016-11-13 paper.tex
reinstating and editing equational theory

\newcommand{\stenReflS}[1]{\term{point} \, (\term{dim=}#1)}
\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\irreflS}{\texttt{nonpoint}}

@@

\begin{definition}[Consistency, informal]
  A stencil specification defines a \emph{region} associated to an
  array variable $a$ against which a stencil kernel should be
  \emph{consistent}. Informally, the region defined by a specification
  is an $n$-dimensional rectangle crossing the origin.
  Consistency is then a two way requirement.
  On the one hand, all array subscripts on $a$ contributing to the kernel
  should be consistent with the region; on the other, all parts of the
  region should have an index corresponding to that part.
\end{definition}
\todo{Improve the above- haven't quite introduce enough yet to say
more}

@@

\section{A model}

We define a simple set-theoretic denotational model of the semantics
of our stencil specification language. This has a number of purposes:
(1) it serves to explain exactly the meaning of our specifications
(2) it is 
used as the basis of the inference, checking, and synthesis algorithms
(\S~\ref{sec:analysis}); (3) justifies the equational theory and
approximations of the previous section; and (4) can be used to guide
correct implementations. The model domain is over sets of vectors
which we call \emph{index schemes}, which we introduce first before
the model in \S~\ref{subsec:model}.

\begin{definition}
  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
  (\ie{}, the constant integer $c$ in an indexing expression of the
  form $i \pm c$, Def.~\ref{def:neighbour},
  p.~\pageref{def:neighbour}) and $\infty$ represents any constant
  inddex. Throughout $u, v, w$ range over
  index schemes and $u_i$ denotes the $i^{th}$ element of $u$. We
  write $(\domainVal)^n$ for the set of $n$-dimensional schemes.
\end{definition}

Indexing schemes give an abstract
representation of array subscripts in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours which are consistent with the specification.
The relationship between indexing schemes and source language
terms is given precisely by the following function \textsf{schematic} which
maps syntactic index expressions to indexing schemes. This
is used in inference and checking.

@@

The partial function $\schemeN$, also parameterised by a set of
induction variables $I$, uses the above operation $\schemeOne$ to
 maps from the syntax of indices to index schemes, \ie{} values of 
$\domainVal^n$. It is defined on the syntax of index
expressions $\bar{e} = (e_1, ..., e_n)$ as:
\begin{align*}
\schemeN(e_1, ..., e_n) & =
[\schemeOne(e_1) ... \schemeOne(e_n) ]
\quad \textit{iff} \;\; \forall x . \schemeOne(e_x) \neq \bot
\end{align*}
%
Thus, for an indexing expression, if each component is a neighbourhood
index, then $\schemeN$ maps it to an indexing scheme.
\end{definition}

@@

\begin{example}
Consider the following three indexing schemes
corresponding to array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$:
%
\begin{align*}
\begin{array}{ll}
\schemeN{(i, j+1)} & =\vtwoh{0}{1} \\
\schemeN{(1, j+1)} & = \vtwoh{\infty}{1} \\
\schemeN{(3, j)} & = \vtwoh{\infty}{0}
\end{array}
\end{align*}
There is one notion of generalisation between these
schemes, namely that $\vtwoh{0}{1} \preceq_2 \vtwoh{\infty}{1}$.
At the level of syntax, this translates to saying
$(i, j+1)$ can be instantiated (specialised) to $(3, j+1)$.

As an example of undefined \textsf{schematic},
$\textsf{schematic}(i*2, j) = \bot$ as its first index is a
non-neighbour relative index, which is outside the scope of our
specification language.
\end{example}

@@

\begin{definition}[Quotienting schemes]
A scheme $u$ can be quotiented by a set of dimension identifies
by the operator $/_{n,I}$ where $I$ is a set of dimension
identifiers drawn from $\mathbb{N}_{\leq n}$ where 
\begin{equation*}
u/_{n,I} = [ ]
\end{equation*}
\end{definition}

@@

Our model is given by the interpretation function $\interp{-}_n$
(where $n$ is the maximum dimensionality of the specification being
modelled) mapping closed\footnote{That is, we assume there are no
  occurrences of \textit{rvar} in a specification being modelled.  Any
  \emph{open} specification containing region variables can be made
  closed by straightforward syntactic substitution with a (closed)
  \textit{region}.} specifications to sets of
$n$-dimensional indexing schemes. The interpretation is overloaded on
\emph{regions} in Figure~\ref{fig:region-model} and on the top-level
of a specification \textit{spec} in Figure~\ref{fig:spatial-model}.

@@

>>>>

\newcommand{\stenReflS}[1]{\term{pointed} \, (\term{dim=}#1)}
\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\irreflS}{\texttt{np}}

@@

\marginpar{\dnote{Is it worth having this informal consistency here?}}
\begin{definition}[Consistency, informal]
  A stencil specification defines a \emph{region} associated to an
  array variable $a$ against which a stencil kernel should be
  \emph{consistent}. Informally, the region defined by a specification
  is an $n$-dimensional rectangle crossing the origin.
  Consistency is then a two way requirement.
  On the one hand, all array subscripts on $a$ contributing to the kernel
  should be consistent with the region; on the other, all parts of the
  region should have an index corresponding to that part.
\end{definition}

@@

\paragraph{Region declarations and variables}

Region specifications can be assigned to region variables via
a region declaration, which can be used later in another
specification. For example, the last specification 
 example above can be restated as:
%%
\begin{minted}{fortran}
!= region :: r1 = forward(depth=1, dim=1)
!= region :: r2 = forward(depth=1, dim=2)
!= region :: robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is especially useful for common stencil patterns, such as Roberts cross,
as the region can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified 
by \emph{approximation} and \emph{multiplicity} information 
(in the \textit{spec} rule of Fig.~\ref{fig:syntax}).

The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, in all of
the previous examples the \texttt{readOnce} modifier could be added,
\eg{}
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of \emph{linearity} in linear type
systems~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

In some cases, it is useful to give a lower and/or upper bound for a
stencil. This can be done using either the \term{atMost} or
\term{atLeast} modifiers. This is particularly useful in situations
where there is a non-contiguous stencil pattern, which cannot be expressed
precisely in our specification syntax. For example:
%
\begin{minted}{fortran}
!= stencil atLeast, pointed(dim=1)       :: a
!= stencil atMost, forward(depth=4, dim=1) :: a
b(i) = a(i) + a(i+4)
\end{minted}

\section{A model for specifications}
\label{sec:semantics}

We define a simple set-theoretic denotational model of the semantics
of our specification language. This has a number of purposes:
(1) it serves to explain the meaning of our specifications
(2) it is used in the inference and checking algorithms
(Section~\ref{sec:analysis}); (3) justifies an equational theory for
specifications in the next section; and (4) can be used to guide
correct implementations. The model domain is over sets of vectors
which we call \emph{index schemes}.

\begin{definition}
  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
  (\ie{}, the constant integer $c$ in an indexing expression of the
  form $i \pm c$, Def.~\ref{def:neighbour},
  p.~\pageref{def:neighbour}) and $\infty$ represents any constant
  index. Throughout $u, v, w$ range over
  index schemes and $u_i$ denotes the $i^{th}$ element of $u$. We
  write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
\end{definition}

Indexing schemes give an abstract
representation of array subscripts in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours which are consistent with the specification.
The relationship between source language
terms and indexing terms
 is given precisely by the following function \textsf{schematic},
 which forms part of checking and inference (\S\ref{sec:analysis}).

@@

The partial function $\schemeN$, also parameterised by a set of
induction variables $I$, uses the above operation $\schemeOne$ to
 map from the syntax of indices to index schemes, \ie{} values of 
$(\domainVal)^n$. It is defined on the syntax of indices $\bar{e} = (e_1, ..., e_n)$ as:
\begin{align*}
\schemeN(e_1, ..., e_n) & =
[\schemeOne(e_1) ... \schemeOne(e_n) ]
\quad \textit{iff} \;\; \forall x . \schemeOne(e_x) \neq \bot
\end{align*}
%
Thus, if each expression in an index is a neighbourhood or constant
index, then $\schemeN$ maps it to an indexing scheme.
\end{definition}

@@

\begin{example}
Consider the following indexing schemes
corresponding to array indexing terms
$(i, j+1)$, $(1, j+1)$, $(3, j)$ and $(i*2, j)$:
%
\begin{align*}
\begin{array}{ll}
\schemeN{(i, j+1)} & =\vtwoh{0}{1} \\
\schemeN{(1, j+1)} & = \vtwoh{\infty}{1} \\
\schemeN{(3, j)} & = \vtwoh{\infty}{0} \\
\schemeN{(i*2, j)} & = \bot
\end{array}
\end{align*}
There is one generalisation between these
schemes, namely that $\vtwoh{0}{1} \preceq_2 \vtwoh{\infty}{1}$.
At the level of syntax, this translates to saying
$(i, j+1)$ can be instantiated (specialised) to $(1, j+1)$. 
Note the last example is undefined as its first expression is a
non-neighbour relative index.
\end{example}

@@

@@

The model is defined by the interpretation function $\interp{-}_n$
(where $n$ is the maximum dimensionality of the specification being
modelled) mapping closed\footnote{That is, we assume there are no
  occurrences of \textit{rvar} in a specification being modelled.  Any
  \emph{open} specification containing region variables can be made
  closed by straightforward syntactic substitution with a (closed)
  \textit{region}.} specifications to sets of
$n$-dimensional indexing schemes. The interpretation is overloaded on
\emph{regions} in Figure~\ref{fig:region-model} and on the top-level
of a specification \textit{spec} in Figure~\ref{fig:spatial-model}.

@@

\paragraph{Model of \textit{spec}}

Regions are modelled as sets of index
schemes, that is
$\interp{\textit{region}}_n \in \mathcal{P}(\domainVal{}^n)$.
Figure~\ref{fig:spatial-model} defines the top-level model of the
\textit{spec} syntax, which includes the multiplicity and
approximation modifiers. The domain of this interpretation
is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
labelling our set-based model with additional information.

\begin{definition} \textsf{Mult}
and \textsf{Approx} are parametric labelled variant types
with injection function given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
The $\textsf{Approx}$ type corresponds to the presence or absence of
the spatial approximation modifier, with \textsf{exact} when there is
no such modifier and \textsf{lower} and \textsf{upper} for
\term{atLeast} and \term{atMost}. Later, it will be useful to model
pairs of lower and upper bounds which is provided by \textsf{both} 
but which is not used here.% These are used for labelled models, and other components later in.
\label{def:mult-and-approx}
\end{definition}

\begin{figure}
\input{spatial-model.tex}
\caption{\textit{spec} model,
$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
\label{fig:spatial-model}
\vspace{-0.75em}
\end{figure}

\section{Equational theory}
\label{sec:eqs}

Specifications are subject to a set of equations which can be used to
understand when specifications are equivalent. Furthermore,
specifications are subject to a notion of approximation. We outline
the equational theory and notion of approximation here.

\begin{figure}
%\vspace{-0.2em}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.6em]
\begin{array}{rll}
(\textsc{F\, \texttt{+} \,F}) \;\; &
\stenFwdSR{n \, \textsf{max} \, m}{d}{p \vee q} \\
 \equiv \; & \stenFwdSR{n}{d}{p} \; \texttt{+} \; \stenFwdSR{m}{d}{q} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,B}) \;\; &
\stenBwdSR{n \, \textsf{max} \, m}{d}{p \vee q} \\
 \equiv \; & \stenBwdSR{n}{d}{p} \; \texttt{+} \; \stenBwdSR{m}{d}{q} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,C}) \;\; &
\stenCenSR{n \, \textsf{max} \, m}{d}{p \vee q} \\
\equiv \; & \stenCenSR{n}{d}{p} \; \texttt{+} \; \stenCenSR{m}{d}{q} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{p \vee q} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{p} \; \texttt{+} \;
                      \stenFwdSR{m}{d}{q} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,B}) \;\; &
\stenCenSR{n}{d}{p \vee q} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{p} \; \texttt{+} \;
                      \stenBwdSR{m}{d}{q} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,F}) \;\; &
\stenCenSR{n}{d}{p \vee q} \\
\equiv \; & \stenFwdSR{n}{d}{p} \; \texttt{+} \; \stenBwdSR{n}{d}{q}
  \\[0.6em]
%
%
(\textsc{P\, \texttt{+} \,F}) \;\; &
\stenFwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{p} \\[0.6em]
%
%
(\textsc{P\, \texttt{+} \,B}) \;\; &
\stenBwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{p} \\[0.6em]
%
%
(\textsc{P\, \texttt{+} \,C}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{p} \\[0.6em]
%
%
(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                         (R^\irreflS \, \term{*} \, S) \;
\equiv \; R \, \term{*} \, S \\[0.4em]
%
(\textsc{$\ast$Dim}) \;\; & R^d \, \term{$\ast$} \, S^d \equiv R^d
                             \, \term{+} \, S^d \\[0.4em]
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.4em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.4em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)
\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications (\emph{regions})}
\vspace{-0.5em}
\label{fig:equations}
\end{figure}

Figure~\ref{fig:equations} lists the equational theory for
specifications given by the binary relation $\equiv$
on \emph{region} terms. The equations use abbreviations
\term{point}, \term{fwd}, \term{bwd}, \term{cen}
for \term{nonpointed}, \term{forward}, \term{backward},
\term{centered} respectively. Furthermore, the abbreviations include syntactic sugar
for quantifying over the absence or presence of the \irrefl{}
attribute on region constants as a boolean. For example,
\term{fwd} syntactic sugar is defined:
%%
\begin{align*}
\hspace{-0.25em}
& \stenFwdSR{n}{d}{r}
\; := \; \\[-0.4em]
& \qquad \begin{cases}
\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\[-0.1em]
\stenFwd{n}{d}{}  & \;\;\textit{if} \;\,  r
\end{cases} \\
\textit{or} \;\;
& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
\end{align*}
%
and similarly for \term{bwd} and \term{cen}.

Stencil specifications in \textit{spec} are considered
equal when they have the same modifiers and $\equiv$ equates
their regions, \ie{}:
\begin{align*}
(\textit{region} \equiv \textit{region'}) 
 \Rightarrow \quad\;\; & \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
\textit{region} \\[-0.4em]
\quad \equiv \; & \texttt{stencil} \; [\textit{approx},] \;
            [\textit{mult},] \; \textit{region}'
\end{align*}
%

\paragraph{Overlapping regions with +}
The first nine
equations (with labels of the form $\ldots \texttt{+} \ldots$)
explain the notion of \emph{region overlapping} with sum
\term{+}. In the first six, $p$ and $q$ range
over booleans denoting the absence (true) or presence (false)
of \irrefl{}. These are combined via disjunction. For example,
the rule $(\textsc{B} \, \term{+} \,
\textsc{F})$ explains that the sum of \term{backward}
and \term{forward} regions in the same dimension and the same
depth is equivalent to a \term{centered} region of that depth.
If the \term{forward} part is non-pointed ($p =
\textit{false}$) but \term{backward} is pointed ($q =
\textit{true}$) then their sum is equal to a \term{centered}
region with no \irreflS{} modifier by the disjunction ($p \vee q =
\textit{true}$), \ie{}:
%%
\begin{align*}
& \stenCenS{n}{d}  \\[-0.2em]
\equiv \;\; & \stenFwdSR{n}{d}{\irrefl} \; \texttt{+} \; \stenBwdS{n}{d}
\end{align*}
%%
That is, the \term{fwd} region is irreflexive describing code which is missing
a zero-offset neighbourhood index (\eg{}, missing $\texttt{a(i)}$), but the addition of
the \term{bwd} has no irreflexive attribute and thus provides the
missing zero offset index.

The equation (\textsc{irrefl}) provides an annihilation of
\irrefl{} using a further syntactic sugar.
We write $R^{\irreflS}$ to denote a region constant with an
non-pointed attribute, \eg{},
%%
\begin{align*}
\stenFwdS{n}{d}^{\irreflS}  := \,
\stenFwdSR{n}{d}{\irrefl}
\end{align*}
%%
The left-hand side of (\textsc{irrefl}) has a sum of
products of two regions $R$ and $S$, 
where in each component of the sum one of $R$
or $S$ has the irreflexive attribute. These cancel to give
just $R \, \term{*} \, S$.

The equation (\textsc{$\ast$Dim}) uses syntactic sugar
where $R^d$ and $S^d$ denote regions whose dimension is $d$.
Thus, the rule shows that a product of two regions on 
the same dimension is equal to their addition.
